

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Introduction

```{r, fig.align='center', echo = FALSE, fig.alt= "Ethical Data Handling for Cancer Research. Written by: Carrie Wright. Contributed and Edited by: Candace Savonen and Jeff Leek", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.p")
```

## Motivation

One of the key cancer informatics challenges is dealing with and managing the explosion of data from multiple sources. This course is designed to help researchers and investigators to understand the key ethical principles of data management from a privacy, security, usability and discoverability perspective. 

## Target Audience 

This course is intended for researchers (including postdocs and students) with limited to intermediate experience with informatics research. The conceptual material will also be useful for those in management roles who are collecting data and using informatics pipelines.

## Topics Covered

```{r, fig.align='center', echo = FALSE, fig.alt= "Concepts discussed in the Ethical Data Handling for Cancer Research course: Risks of personal identifiable information and personal health information. Tips for HIPAA compliance and methods of de-identification. Cancer data types that are potentially re-identifiable. Major methods of data security and practical everyday tips. Tips about sharing cancer research data, including relevant repositories. General History of medical research ethics. Current and potential concerns for medical research ethics. Strategies to protect and respect research participants.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g256fece0af0_0_0")
```


## Curriculum 

The course will cover the key underlying principles and concepts in data management. It will also cover data security, data privacy, IRB and data access requests, and the ethics of good data management. 

```{r, fig.align='center', echo = FALSE, fig.alt= "Overall Course Learning Objectives. This course will demonstrate how to: 1.Implement best practices for data security, 2.Recognize and work with data that requires extra privacy protection, 3.Share data safely and securely in a variety of contexts, 4. Apply ethical consideration in data management workflows", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.gf5f8818810_1_5")
```
