

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Introduction

```{r, fig.align='center', echo = FALSE, fig.alt= "Ethical Data Handling for Cancer Research. Written by: Carrie Wright. Contributed and Edited by: Candace Savonen and Jeff Leek", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.p")
```

## Motivation

One of the key cancer informatics challenges is dealing with and managing the explosion of data from multiple sources. This course is designed to help researchers and investigators to understand the key ethical principles of data management from a privacy, security, usability and discoverability perspective. 

## Target Audience 

This course is intended for researchers (including postdocs and students) with limited to intermediate experience with informatics research. The conceptual material will also be useful for those in management roles who are collecting data and using informatics pipelines.


```{r for_individuals_who, fig.align='center', echo = FALSE, fig.alt= "This courses is itended for anyone who needs a refresher on ethical data handling, Is new working with new data (esp. clinical or genomic data) and wants to use it appropriately, wants to learn about the emerging and current issues in biomedical ethics", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g256827a0886_0_0")
```


## Topics Covered

```{r topics_covered, fig.align='center', echo = FALSE, fig.alt= "Concepts discussed in the Ethical Data Handling for Cancer Research course: Risks of personal identifiable information and personal health information. Tips for HIPAA compliance and methods of de-identification. Cancer data types that are potentially re-identifiable. Major methods of data security and practical everyday tips. Tips about sharing cancer research data, including relevant repositories. General History of medical research ethics. Current and potential concerns for medical research ethics. Strategies to protect and respect research participants.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g256fece0af0_0_0")
```


## Curriculum 

The course will cover  key underlying principles and concepts in ethical data handling. 

```{r learning_objectives, fig.align='center', echo = FALSE, fig.alt= "Overall Course Learning Objectives. This course will demonstrate how to: 1.Recognize and work with data that requires extra privacy protection, 2.Implement best practices for data security, 3.Share data safely and securely in a variety of contexts, 4. Apply ethical consideration in data management workflows, 5. Recognize current ethical concerns regarding data for research",  out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.gf5f8818810_1_5")
```

It will also cover how best practices for data security, data privacy, and data sharing are critical for responsible data management.

```{r, fig.align='center', echo = FALSE, fig.alt= "Ethical data practices include: Sharing data safely and openly to help further the research field. Keeping sensitive data private and hidden in order to protect patients from harm. Securing data so it is not exposed and cannot be manipulated or found by others who are not permitted to access it.", out.width= "100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g11ef3a2f2bf_0_0")
```

