

```{r, include = FALSE}
leanbuild::set_knitr_image_path()
```

# Data Sharing

In this chapter we will discuss the importance of data sharing, best practices for data sharing, places where you can store your data, as well as methods to share data in contexts in which you might have thought sharing was not possible!

```{r, fig.align='center', echo = FALSE, fig.alt= "Learning Objectives: 1.Define the term data security, 2. Describe major data security strategies, 3.Explain the basics of how encryption works, 4.Be aware of how data security practices may or may not be a part of your workflow", out.width="100%"}
leanbuild::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g1016753ce66_0_24")
```

## Data sharing is important!

Sharing data is critical for optimizing the advancement of scientific understanding. Now that labs all over the world are producing massive amounts of data, there are many discoveries that can be made by just using this existing data.

This is so important, that starting in January, 2023 the NIH will require specific sharing practices for data management and sharing. See the announcement [here](https://grants.nih.gov/grants/guide/notice-files/NOT-OD-21-013.html).

## Put your data in repository

### Genomic Data Repositories

- [National Center for Biotechnology Information (NCBI)](https://www.ncbi.nlm.nih.gov/)
  - For microarray: [GEO Gene Expression Omnibus (GEO)](https://www.ncbi.nlm.nih.gov/geo/info/submission.html)
  - For RNA-seq: [SRA (Sequencing Read Archive)](https://www.ncbi.nlm.nih.gov/sra/docs/submit/)
- [European Molecular Biology Laboratory-European Bioinformatics Institute (EMBL-EBI)](https://www.ebi.ac.uk/submission/)
- [International Nucleotide Sequence Databases—DNA Data Bank of Japan (DDBJ)](https://www.ddbj.nig.ac.jp/ddbj/submission-e.html)

### Imaging data repositories

- [Imaging data resource](https://idr.openmicroscopy.org/)
AVOCADO: Others?

### Repositories for journal articles

For manuscripts or large datasets that are of atypical format, using one of these repositories is a good idea.
The journal you submit to may have a recommendation of one over another. If not, you might end up having a preference.

- [CyVerse Data Commons Repository](https://cyverse.org/data-commons)
- [Data Dryad](https://datadryad.org/stash)
- [FigShare](https://help.figshare.com/article/how-to-upload-and-publish-your-data)
- [ZENODO](https://help.zenodo.org/)

### Small datasets

Data sets that are small and atypical format can be published as supplementary files as a part of a manuscript.

## Data Submission tips

### Have someone else review your code and data!

There are so many little details that go into your data and projects. Those details can easily lead to typos and errors upon data submission and also can lead to confusion when others (or your future self) are attempting to use that data.
The best way to test if your data project is useable is to have someone else (who has not prepared the data) is able to make sense of it.

### Use consistent and clear names

- Make sure that sample and data IDs used are consistent across the project - make sure to include a metadata file that describes in detail your samples in a way that is clear without any prior knowledge of the project.
- Sample and data IDs should keep with standard formatting otherwise known in the field.
- Features names should avoid using genomic coordinates as these may change with new genome versions.

## Make your project reproducible

Reproducible projects are able to be re-run by others to obtain the same results.

The main requirements for a reproducible project are:
- The data can be freely obtained from a repository (this maybe summarized data for the purposes of data privacy).
- The code can be freely obtained from GitHub (or another similar repository).
- The software versions used to obtain the results are made clear by documentation or providing a Docker container.
- The code and data are well described and organized with a system that is consistent.

For more details on how to make data and code reproducible tips, see our [Intro to Reproducibility](https://www.itcrtraining.org/courses#h.ugabyqq1bigx) course.
