

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Data Ethics

Now that we have covered the basics of data management, we will take a moment to consider and reflect on the implications of our use and sharing of data.

```{r, fig.align='center', echo = FALSE, fig.alt= "Learning Objectives: 1.Define the term data security, 2. Describe major data security strategies, 3.Explain the basics of how encryption works, 4.Be aware of how data security practices may or may not be a part of your workflow", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g1016753ce66_0_24")
````

Data ethics involves the consideration of the aspects that we have already described about data security, privacy, and sharing, in terms of how our research can ultimately impact research participants and other individuals. Other general principles have to do with research integrity, consideration for social justice, and the transparency of our work. This requires thoughtfulness throughout the planning and research process, as well as mindfulness for what happens after the research is complete and published. Researchers need to consider how others might use our data, code, and results in the future [@ipworth_ethics_2017; @teoli_informatics_2021].








