

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Data Ethics

Now that we have covered the basics of data management, we will take a moment to consider and reflect on the implications of our use and sharing of data.

```{r, fig.align='center', echo = FALSE, fig.alt= "Learning Objectives: 1.Define the term data security, 2. Describe major data security strategies, 3.Explain the basics of how encryption works, 4.Be aware of how data security practices may or may not be a part of your workflow", out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1SRokLaGAc2hiwJSN26FHE0ZEEhPr3KQdyMICic8kAcs/edit#slide=id.g1016753ce66_0_24")
````

Data ethics involves the consideration of the aspects that we have already described about data security, privacy, and sharing, in terms of how our research can ultimately impact research participants and other individuals. Other general principles have to do with research integrity, consideration for social justice, and the transparency of our work. This requires thoughtfulness throughout the planning and research process, as well as mindfulness for what happens after the research is complete and published. Researchers need to consider how others might use their data, code, and results in the future [@ipworth_ethics_2017; @teoli_informatics_2021]. 

While data sharing can result in wonderful opportunities for secondary analysis, we need to also consider some of the harm that could me caused by sharing our data and make sure that we do it mindfully. With more advanced forms of genomic and imaging technology, we have much more information about the subjects we are using, and thus the risk for others to identify the subjects in our studies is much higher than it used to be [@byrd_responsible_2020]. 


## Data Ethics History

To have an understanding of current theories about how to best deal with our research ethic conundrums it is helpful to be aware of the history of biomedical research in general.

Most regulations of biomedical research stems from historical ill treatment of research participants. This has taken many forms from outright ill intent, to much more well-intended but neglectful research practices often due to a lack of awareness of the potential consequences. This has been especially difficult more recently as our potential to create and share data has dramatically expanded.

Avocado - want to provide examples for databanks gone wrong, such as the use of data for commercial purposes. Could go back further too... to  HIPAA in 1996 and 1957 Code of Medical Ethics.


## General Ethics Code

- informed consent etc.

## Current Ethical Issues

There are several issues that researchers and research participants, and really all individuals engaging in health care face. 


### Consent for Data Reuse

One major current ethical issue that we face now, is the consequences of the reuse of shared data. As we have described, there are major benefits of sharing data. It can allow researchers to really maximize their efforts. However, there are also negative potential consequences as well.


Previous management strategies for informed consent originate from research that predates the large scale data sharing that we now use today. In those cases, informed consent was a bit more straight forward to achieve. Now that data is often reused more often, it is often less obvious how data will be used for research purposes in the future, thus it is less obvious how to inform potential research participants what participation really means. Ideally we want to protect participants and family, while also maximizing the research potential of useful data. So how do we do this?

Although ethical guidelines about this type of consent are evolving as research and technology evolve, here are some methods for consent that some ethicist suggest as described in @mckeown_consent_2021.

1) blanket consent - 
2) broad consent -
3) dynamic consent -
4) meta consent -

hmm or how it is described here:https://www.nature.com/articles/s41576-020-0257-5.pdf

### Finding Artifacts

In some cases researchers will determine aspects about the potential health or genetic risk of an individual as an artifact of performing other research. This leads to the question of if those individuals should be informed about these findings.




### Protection of Vulnerable Populations


### Equity of Health Care


### Data Security and Privacy in the Age of AI

avocado... possibly introduce: differential privacy https://mccourt.georgetown.edu/research/the-massive-data-institute/resources/dp-resources/









